---
title: "Working with a dataset"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

## Overview

We will install a package containing a dataset and use it to practice some 
exploratory data analysis methods.

### Install the gapminder package

Run the command `install.packages("gapminder")`, followed by 
```{r, warning=FALSE}
library(gapminder)
```

### Store and subset the data

Run `gapminder`.  The output will look a little strange.
Let's get our bearings.

First, store the dataset in your environment with
```{r}
mygapminder <- gapminder
```

This will make the dataset more accessible.

Then, use some commands that we know to get a sense of what's going on:
`str(mygapminder)` and `summary(mygapminder)`.

Use `unique(mygapminder$country)` to examine a list of all the unique values that
`mygapminder$country` can take on.

Let's say we want to focus our analysis on a specific country (intra-national
vs. inter-national). We need to create a subset of the data with only data from
the desired country.

We'll use the `sample()` command to randomly select a country to analyze.
`sample(unique(mygapminder$country), 1)`.  We just told R to take a sample of 1
from the result of `unique()`.

That was the lazy and messy way.  We should do:

```{r, results='hide'}
countries <- unique(mygapminder$country)
sample(countries, 1)
```

This is easier to read, and we have a `countries` object to use later. 

Let's say R chose Iceland.  We can run 
`subset(mygapminder, country == "Iceland")` to return only the data for records 
where country is equal to "Iceland".