---
title: "Basic Operations and Data Types in R"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

## Overview
In this session, we will introduce some basic operations (doing things) and data
types (things you do things to) in R.

- intro to data types
- vectors
- arrow assignment
- basic math functions
- dataframes
- summary functions

If you haven't already done so, install R and RStudio in 
[Part 1](https://timothytsai.github.io/CI670/part1_install_r.html)

### Introduction to data types
There are 5 basic types of data (classes which dictate what kinds of values a 
variable can take) in R:
- logical
- integer
- numeric
- complex
- character

There are many data structures in R (ways of storing and organizing data):
- vector
- list
- matrix
- data frame
- factors
- tables

We will focus on **vectors** and **data frames**.

### Vectors

Vectors are the most basic data structure in R and you will deal with vectors
when using functions and writing your own functions.

```{r}
age.years <- c(25, 40, 77)
```

By running the above line of code, you just created a *numeric* vector using the
`c()` function and stored it as a vector named `age.years`.  

A vector is an ordered sequence of data elements of the same type.  You can have
vectors of each of the data types we mentioned above.

We use an arrow `<-` to assign elements on the **right** to the object on 
the **left**.

`age.years` in our example above is a numeric vector with 3 elements.  If we
want to output the second element in the vector, we run: `age.years[2]`, 
returning `r age.years[2]`.

Vectors are useful because you can perform *vectorized operations*.  A simple
illustration of this is:

```{r}
age.years + 1
```

In the line of code above, we have added 1 to each element of the vector.  This
type of operation is a significant advantage of R.  In some other languages,
you might have to write a loop to add one to each element sequentially until the
loop reaches the end of the object.  R can do this simultaneously to each
element, which is generally more efficient.

#### Important Vector properties

When you encounter a vector in the wild, you should find out two key properties
of the vector before working with it: type and length.

`typeof(age.years)` will tell you that `age.years` is of type "double", a type
of numeric format (see `?double` for possible boring details).

`length(age.years)` tells you that `age.years` has 3 elements.

### Data frames

If we were to create three vectors of equal length and join them together, we
would create an object with 3 rows and 3 columns.

```{r}
age <- age.years
gender <- c("M", "F", "F")
smoking <- c(TRUE, TRUE, FALSE)
```

Above we have just created three vectors of different types but the same length
(3).

We will now use `data.frame()` to join them together

```{r}
mydataframe <- data.frame(age, gender, smoking)
```

We have just created a data frame called "mydataframe" with 3 observations and 3
variables - age, gender, and smoking.

A dataframe is the fundamental method for storing tabular data in R.  You can 
think of it as a list of vectors, or a rectangular list.  When creating a
dataframe from one or more vectors, each vector must be of the same length.

A dataframe has columns and rows which may be modified.

```{r}
colnames(mydataframe)
row.names(mydataframe)
```

#### Indexing a dataframe

When we want to examine an element of an object - in this case, a dataframe, we
need to *index* the desired part of the object.  Like finding a book (a form of
physical media) in a library (a place where books are stored), there is a
logical system that R uses.

The important thing to remember is Row, Column.

```{r}
mydataframe[2,3]
```

Above we have returned the element in the 2nd Row, 3rd Column.  You can verify
this by examining `View(mydataframe)`.

If we run `mydataframe[2,]`, we return the 2nd row and `mydataframe[, 3]` returns
the 3rd column.  Note that you need to keep the comma and put it in the 
appropriate location.

Once you know the column names of the dataframe, you can also return that column
by using the `$` symbol - `mydataframe$age`, for example.