---
title: "Writing a function"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

## Overview

In this module, we'll learn about functions and write a simple function to plot
positive predictive value (PPV) given sensitivity, specificity, and prevalence.

### Functions

Functions in R are a type of object which at the fundamental level, accept an 
input and return an output.  Contrast this to the mathematical definition of a
function, where each input is related to exactly one output.

A function is a set of instructions that you want to use repeatedly or are so
complex that they are best contained in a single object (as opposed to multiple
lines of code in a script).

#### User-defined functions

This module will focus on user-defined functions, or functions that are 
constructed by the user as opposed to provided by R or an R package.

There are many reasons why you might want to write your own function:

- to avoid repeating chunks of code that do the same thing

- to share a piece of code 

- to automate tasks

By avoiding repeated chunks of code, we are also improving the readability of
the code and reducing the likelihood of errors associated with copying and
pasting.  Improved readability of code means that it is easier to spot errors.
It is also easier to make a correction or update to a single function compared 
to multiple nearly identical pieces of code.  

### Example

In class, we calculated Positive Predictive Value using the following piece of
code:

```{r, eval = FALSE}
ppv <- (sensitivity * prevalence) /
    ((sensitivity * prevalence) + (1 - specificity) * (1 - prevalence))
```

We then calculated different PPVs using the code below, repeated for different
values of sens and spec, saving the new PPVs to objects names ppv1 ... ppv4

```{r, eval = FALSE}
prev <- seq(.1, .4, .025)
sens <- .88
spec <- .6
ppv <- sens * prev / ((sens*prev) + (1-spec)*(1-prev))
```

Finally, we constructed a plot of prevalence and ppv using the below:

```{r, eval = FALSE}
plot(prev, ppv, las = 1, xlim = c(0, 0.5), ylim = c(0, 1))
lines(prev, ppv4, color="red", type="b", pch=16)
```

We can simplify this process using the following function:

```{r, eval = FALSE}
ppv.plot <- function(prevalence, sensitivity, specificity){
ppv <- (sensitivity * prevalence) /
    ((sensitivity * prevalence) + (1 - specificity) * (1 - prevalence))
  plot.out <- plot(prevalence, ppv, las = 1, xlim = c(0, 0.5), ylim = c(0, 1))
 return(plot.out)
}
```